Nonparametric Clustering from Energy Statistics

### 2017-06-21 qcqp_clustering/draft/energy_clustering.tex
working draft

\begin{abstract}
Energy statistics provides a nonparametric test for equality of distributions.
It was proposed by 
Sz\' ekely in the 80's,
inspired by Newton's potential energy between massive bodies. The idea
is to associate a statistical potential energy to observations such that 
minimum energy is achieved under the null hypothesis of equality of 
distributions. This 
was further generalized for probability 
distributions on arbitrary metric spaces,
and more recently, a connection with kernels in RKHS was established.
Nevertheless, although extensively used by the statistics community, it was
not incorporated in machine learning problems.
In this paper, we consider the problem of clustering data based
energy statistics theory. 
We provide a precise mathematical formulation, obtaining
a quadratically constrained optimization problem (QCQP). We show that
this optimization problem is equivalent to kernel $k$-means
optimization problem, however,
energy statistics is able to fix the kernel choice. 
This method is nonparametric, and if prior information is available
it can be easily incorporated in the kernel construction.
We propose an algorithm
to find local optimizers of this QCQP problem based on the energy cost
of moving points to different clusters. We then compare this algorithm with
kernel $k$-means, standard $k$-means, and GMM. We test our method
on synthetic and real data experiments.
Our results show that energy statistics based clustering outperforms
these most used clustering algorithms.
\end{abstract}

>>>>

\begin{abstract}
Energy statistics provides a nonparametric test for equality of distributions.
It was proposed by 
Sz\' ekely in the 80's,
inspired by Newton's potential energy between massive bodies. The idea
is to associate a statistical potential energy to observations such that 
minimum energy is achieved under the null hypothesis of equality of 
distributions. This 
was further generalized for probability 
distributions on arbitrary metric spaces,
and more recently, a connection with kernels in RKHS was established.
Nevertheless, although extensively used by the statistics community, it was
not incorporated in machine learning problems.
In this paper, we consider the problem of clustering data based
energy statistics theory. 
We provide a precise mathematical formulation, obtaining
a quadratically constrained quadratic optimization problem (QCQP). 
We show that
this is equivalent to kernel $k$-means
optimization problem, however,
energy statistics is able to fix the kernel choice. 
This method is nonparametric, and if prior information is available
it can be easily incorporated in the kernel construction.
We propose an algorithm
to find local optimizers of this QCQP problem based on the energy cost
of moving points to different clusters. We then compare this algorithm with
kernel $k$-means, standard $k$-means, and GMM. We test our method
on synthetic and real data experiments.
Our results show that energy statistics based clustering outperforms
these most used clustering algorithms.
\end{abstract}
